Overrelaxed Sinkhorn--Knopp Algorithm for Regularized Optimal Transport

### 2017-10-20 paper.tex
Relecture Alexis

\begin{abstract}
This article describes a method for quickly computing the solution to the regularized optimal transport problem. It generalizes and improves upon the widely-used iterative Bregman projections algorithm (or Sinkhorn--Knopp algorithm). 
The idea is to overrelax the Bregman projection operators, allowing for faster convergence. In practice this corresponds to elevating the diagonal scaling factors to a given power, at each step of the algorithm.
We propose a simple method for establishing global convergence by ensuring the decrease of a Lyapunov function at each step.
An adaptive choice of overrelaxation parameter based on the Lyapunov function is constructed.
We also suggest a heuristic to choose a suitable asymptotic over-relaxation parameter, based on a local convergence analysis. Our numerical experiments show a gain in convergence speed by an order of magnitude in certain regimes.
\end{abstract}

@@

The computation of optimal transport between two data relies on the estimation of an optimal transport matrix, the entries of which represent the quantity of mass transported between  data locations. 
Regularization of optimal transport with strictly convex regularizers \cite{cuturi13, dessein2016}  nevertheless involves a spreading of the mass. Hence, for particular purposes such as color interpolation \cite{Rabin2014} or gradient flow \cite{2016arXiv160705816C}, it is  necessary  to consider very small regularization of the problem.
In this setting,  the regularized transport problem can be ill-conditioned and Sinkhorn's algorithm converges slowly. This is the issue  we want to tackle here.
Before going into further details, we now briefly introduce the main notations and concepts used all along this article.

@@

In \cite{cuturi13}, it has been proposed to regularize this problem by adding a strictly convex entropy regularization:
\begin{equation}\label{ROT}
\min_{\gamma\in\Ccal_1\cap \Ccal_2\cap \IR^{n_1 n_2}_{+}}K^\epsilon(\gamma) \coloneqq \scal{c}{\gamma} 
+ \epsilon \KL(\gamma,\mathbf{1})
,\end{equation}
with $\epsilon>0$, $\mathbf{1}$ is the matrix of size $n_1\times n_2$ full of ones and the Kullback-Leibler divergence is
\begin{equation}\label{KL}
\KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}
\end{equation}
with the convention $0 \log 0= 0$. It was shown in \cite{benamou15}  that the regularized optimal transport matrix $\gamma^*$, which is the unique minimizer of problem \eqref{ROT},  is the Bregman projection of $\gamma^0 = e^{-c/\epsilon}$ (here and in the sequel, exponentiation is meant entrywise) onto $\Ccal_1 \cap \Ccal_2$:
\begin{equation}\label{eq:reg_ot_pb}
\gamma^* = \argmin_{\Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma)= P_{\Ccal_1 \cap \Ccal_2} (e^{-c/\epsilon}),
\end{equation}
where $P_{\Ccal}$ is the  Bregman projection onto $\Ccal$ defined as
\[
P_{\Ccal}(\xi) \coloneqq \argmin_{\gamma \in \Ccal} \KL(\gamma,\xi).
\]

\subsection{Sinkhorn--Knopp algorithm}
Iterative Bregman projections onto $\Ccal_1$ and $\Ccal_2$ converge to a point in the intersection $\Ccal_1 \cap \Ccal_2$ \cite{bregman67}. Hence, the so-called Sinkhorn--Knopp algorithm (SK) \cite{sinkhorn64} that performs alternate Bregman projections, can be considered to compute the regularized  transport matrix:
\begin{align*}
\gamma^0 &= e^{-c/\epsilon} &
\gamma^{l+1} = P_{\Ccal_2}(P_{\Ccal_1}(\gamma^l)),
\end{align*}
and we have 
$\lim_{l\rightarrow +\infty} \gamma^l = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0) = \gamma^*.$
%
In the discrete setting, these projections correspond to diagonal scalings of the input:
\begin{align}\label{scaling}
P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &\text{with}\quad
a &=  {\mu^1}\oslash{A_1 \gamma} \\
P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &\text{with}\quad
b &= {\mu^2}\oslash{A_2 \gamma}\nonumber
\end{align}
where $\oslash$ is the pointwise division. 
To compute numerically the solution one simply has to store $(a^l, b^l)\in\IR^{n_1}\times \IR^{n_2}$ and to iterate
\begin{align*}
a^{l+1} &= {\mu^1}\oslash{\gamma^0 b^l} &
b^{l+1} &= {\mu^2}\oslash{^t \gamma^0 a^{l+1}} .
\end{align*}
We then have $\gamma^l = \diag(a^l) \gamma^0 \diag(b^l).$

Another way to interpret the SK algorithm is as an alternate maximization algorithm on the dual of the regularized optimal transport problem. The dual problem of \eqref{ROT} is
\begin{equation}\label{DROT}
\max_{\substack{\alpha\in \IR^n\\\beta\in \IR^m}}\; E(\alpha,\beta) \coloneqq \langle \alpha,\mu^1\rangle+\langle \beta,\mu^2\rangle-\epsilon\sum_{i,j}e^{(\alpha_i+\beta_j-c_{i,j})/\epsilon}.
\end{equation}
The function $E$ is concave, continuously differentiable and admits a maximizer, so alternate maximization converges and we recover, for $a_i=e^{\alpha_i/\epsilon}$, $b_j=e^{\beta_j/\epsilon}$ and $\gamma^0_{i,j}=e^{-c_{i,j}/\epsilon}$:
\begin{align*}
a^{l+1} &= {\mu^1}\oslash{\gamma^0 b^l} &
b^{l+1} &= {\mu^2}\oslash{^t \gamma^0 a^{l+1}} 
\end{align*}

@@

As illustrated in Figure \ref{alternate_projections} (a-b), the SK algorithm, that  performs alternate Bregman projections onto the affine sets $\Ccal_1$ and $\Ccal_2$, can be very slow when $\epsilon\to 0$. The idea developped in this paper is to perform overrelaxed projections in order to accelerate the process, as displayed in Figure \ref{alternate_projections} (c).

\begin{figure}[ht!]
\centering
\begin{minipage}[b]{.33\linewidth}
   \centering
   \input{schema_a.tex}
                    \subcaption{}\label{fig:schema_a}
\end{minipage}%
\begin{minipage}[b]{.33\linewidth}
   \centering
   \input{schema_b.tex}
                    \subcaption{}\label{fig:schema_b}
\end{minipage}%
\begin{minipage}[b]{.33\linewidth}
   \centering
   \input{schema_c.tex}
                    \subcaption{}\label{fig:schema_c}
\end{minipage}%
\caption{\label{alternate_projections} The trajectory of $\gamma^l$ given by the SK algorithm is illustrated for decreasing values of $\epsilon$ in (a) and (b). Overrelaxed projection typically accelerate the convergence rate.}
\end{figure}

\iffalse
\begin{figure}[ht!]
\begin{center}
\begin{tabular}{ccc}
\input{schema_a.tex}&
\input{schema_b.tex}&
\input{schema_c.tex}\\
(a)&(b)&(c)
\end{tabular}
\caption{\label{alternate_projections} The trajectory of $\gamma^l$ given by the SK algorithm is illustrated for decreasing values of $\epsilon$ in (a) and (b). In order to accelerate the convergence rate, over-relaxed projections are here considered (c).}
\end{center}
\end{figure}
\fi

@@

We define the $\omega$-overrelaxed projection operator $P^\omega_{\Ccal_k}$ as
\begin{equation}\label{eq:def_or_proj}
\log P^\omega_{\Ccal_k}(\gamma) = (1-\omega) \log \gamma + \omega \log P_{\Ccal_k}(\gamma),
\end{equation}
where the logarithm is taken coordinate-wise. Note that $P_{\Ccal_k}^0$ is the identity, $P_{\Ccal_k}^1 = P_{\Ccal_k}$ is the standard Bregman projection and $P_{\Ccal_k}^2$ is an involution (in particular because $\Ccal_k$ is an affine subspace). A naive algorithm would then consist in iteratively applying $P^\omega_{\Ccal_2}\circ P^\omega_{\Ccal_1}$ for some choice of $\omega$. This algorithm, while it may behave well in practice, has no convergence guarantee---it sometimes does not convergence even for reasonable values of $\omega$. Our goal in this section is to modify this algorithm to make it robust and garantee convergence.

Duality gives another point of view on the iterative overrelaxed Bregman projections. They indeed correspond to a successive overelaxation (SOR) algorithm on the dual objective $E$. This can be seen by using the relationships gievn after equation~\eqref{DROT}:
\begin{align}\label{SORdual}
\alpha^{\ell+1} &= (1-\omega)\alpha^{\ell} + \omega \arg \max_\alpha E(\alpha,\beta^{\ell})\\
\beta^{\ell + 1}&=(1-\omega) \beta^{\ell} + \omega \arg \max_\beta E(\alpha^{\ell+1},\beta).
\end{align}

{\color{blue} \begin{proposition} 
State here that the fixed points of   $\gamma^{l+1} =P^\theta_{\Ccal_2}(P^\theta_{\Ccal_1}(\gamma^l))$ are the same than $\gamma^{l+1} =P_{\Ccal_2}(P_{\Ccal_1}(\gamma^l))$? lenaic: what is exactly needed? I can think of it tomorow.
\end{proposition}}
\subsection{Lyapunov function}
Convergence of the successive overrelaxed projection cannot be guaranteed in general. In order to derive a robust algorithm with provable convergence, we introduce the Lyapunov function 
\begin{equation}\label{eq:lyapunov_function}
F(\gamma) = \KL(\gamma^*, \gamma),
\end{equation}
where $\gamma^*$ denotes the solution of the regularized OT problem. We will use this function to enforce the strict descent criterion $F(\gamma^{k+1}) < F(\gamma^k)$ as long as the process has not converged.
%
The choice of \eqref{eq:lyapunov_function} as a Lyapunov function is of course related to the fact that Bregman projections are used throughout the algorithm. Further, we will show (Lemma \ref{lemma:lyapunov_decrease}) that its decrease is very easy to compute and this descent criterion still allows enough freedom in the choice of the overrelaxation parameter.

@@

\iffalse
\begin{proof}
	First note that K is closed in $\IR_{+*}^{n_1 n_2}$, by continuity of $F$.
	For $\gamma \in K$, one has:
	\begin{align*}
	F(\gamma) &= \scal{\gamma^*}{\log \gamma^* - 1} + \scal{\gamma}{1} - \scal{\gamma^*}{\log \gamma}\\
	\sum_{i,j} \gamma_{i,j} - \gamma^*_{i,j} \log \gamma_{i,j} &\le M - \scal{\gamma^*}{\log \gamma^* - 1} .
	\end{align*}
	By concavity, the logarithm admits a simple upper bound: $\log \gamma_{i,j} \le \frac{\gamma_{i,j}}{e}$. Moreover, since $\gamma^*$ is a solution of the problem, each of its coordinates is at most 1.
	\begin{align*}
	\sum_{i,j} \gamma_{i,j} &\le \frac{M - \scal{\gamma^*}{\log \gamma^*-1}}{1-\frac{1}{e}} =: M'.
	\end{align*}
	Therefore each coordinate $\gamma_{i,j}$ is bounded above by $M'$.

	Let us show that the coordinates are bounded below by a strictly positive number; from which it can be deduced that $K$ is bounded and closed in $\IR^{n_1 n_2}$, and therefore compact.
	\begin{align*}
	F(\gamma)
	&\ge \scal{\gamma^*}{\log \gamma^*} - \sum_{i,j} \gamma^*_{i,j} \log(\gamma_{i,j}) - 1\\
	&\ge \scal{\gamma^*}{\log \gamma^*} - \gamma^*_{i_0,j_0} \log(\gamma_{i_0,j_0}) - 1 - \log M',
	\end{align*}
	where the last equation derives from $\log(y_{i,j}) \le \log M'$, and $\sum_{(i,j)\neq(i_0,j_0)} \gamma^*_{i,j} \le 1$.
	Finally:
	\[
	\log(\gamma_{i_0,j_0}) \ge \frac{\scal{\gamma^*}{\log \gamma^*} - M - \log M' - 1}{\gamma^*_{i_0,j_0}},
	\]
	and $K$ is thus compact.
\end{proof}
\fi

@@

\subsection{Proposed algorithm}
We first give a general convergence result that later serves as a basis to design an explicit algorithm.
\begin{theorem}\label{thm:algo}
Let $\theta_1$ and $\theta_2$ be two continuous functions of $\gamma$ such that
	\begin{equation}\label{eq:cond_theta_k}
	\forall \gamma \in \IR_{+*}^{n_1 n_2},\quad
	F(P_{\Ccal_k}^{\theta_k(\gamma)}(\gamma)) \le F(\gamma) ,
	\end{equation}
	where the inequality is strict whenever $\gamma \notin \Ccal_k$.
	Consider the sequence defined by $\gamma^0 = e^{-c/\epsilon}$ and
	\begin{align*}
	\tilde{\gamma}^{l+1} &= P_{\Ccal_1}^{\theta_1(\gamma^l)}(\gamma^l) \\
	\gamma^{l+1} &= P_{\Ccal_2}^{\theta_2(\tilde{\gamma}^{l+1})}(\tilde{\gamma}^{l+1}).
	\end{align*}
	Then the sequence $(\gamma^l)$ converges to $\gamma^*$.
\end{theorem}

\begin{lemma}[\cite{cuturi13}]
	\label{lemma:trivial_intersection}
	Let us take $\gamma^0$ in $\IR_{+*}^{n_1 n_2}$,
	and denote
	\[
	S = \left\{
	\diag(a) \gamma^0 \diag(b),\quad
	(a,b) \in \IR_{+*}^{n_1 + n_2}
	\right\}
	\]
	the set of matrices that are diagonally similar to $\gamma^0$.
	Then the set $S \cap \Ccal_1 \cap \Ccal_2$ contains exactly one element $\gamma^* = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0)$.
\end{lemma}

\begin{proof}[Proof of the theorem]
	First of all, notice that the operators $P_{\Ccal_k}^\theta$ apply a scaling to lines or columns of matrices. All $(\gamma^l)$ are thus diagonally similar to $\gamma^0$:
	\[
	\forall l\ge0,\quad \gamma^l \in S
	\]
	By construction of the functions $\theta_k$, the sequence of values of the Lyapunov function $(F(\gamma^l))$ is non-increasing. Hence $(\gamma^l)$ is precompact.
	%
	If $\xi$ is a cluster point of $(\gamma^l)$, let us define
	\begin{align*}
	\tilde{\xi} &= P_{\Ccal_1}^{\theta_1(\xi)}(\xi) \\
	\xi' &= P_{\Ccal_2}^{\theta_2(\tilde{\xi})}(\tilde{\xi}).
	\end{align*}
	Then by continuity of the applications, $F(\xi) = F(\tilde{\xi}) = F(\xi')$.
	From the hypothesis made on $\theta_1$ and $\theta_2$, it can be deduced that $\xi$ is in $\Ccal_1$ and that $\tilde{\xi}$ is in $\Ccal_2$. Therefore $\xi' = \tilde{\xi} = \xi$ is in the intersection $\Ccal_1 \cap \Ccal_2$.
	By Lemma \ref{lemma:trivial_intersection}, $\xi = \gamma^*$, and the whole sequence $(\gamma^l)$ converges to the solution.
\end{proof}

@@

To construct explicitly functions $\theta_k$ as needed in Theorem~\ref{thm:algo}, we need the following lemma.
\begin{lemma}\label{lemma:F_P_theta}
	Let $1\le \theta < \omega$. Then, for any $\gamma \in \IR_{+*}^{nm}$, one has
	\begin{equation}\label{eq:F_P_theta}
	F(P^\theta_{\Ccal_k}(\gamma)) \le F(P^\omega_{\Ccal_k}(\gamma)).
	\end{equation}
	Moreover, equality occurs if and only if $\gamma \in \Ccal_k$.
\end{lemma}
\begin{proof}
	Thanks to lemma \ref{lemma:lyapunov_decrease}, one knows that
	\[
	F(P^\theta_{\Ccal_k}(\gamma)) - F(P^\omega_{\Ccal_k}(\gamma))
	= \scal{\mu^k}{(\varphi_\omega - \varphi_\theta) \left( \frac{A_k \gamma}{\mu^k} \right) } .
	\]
	Moreover, the function that maps $t \in [1,\infty)$ to $\varphi_t(x)$ is non-increasing:
	$\partial_t \varphi_t(x) =  (x^{1-t} - 1)\log x.$
	For $x\neq 1$, it is even strictly decreasing.
	Thus inequality \eqref{eq:F_P_theta} is valid, with equality \emph{iff} $A_k \gamma = \mu$.
\end{proof}

@@

\begin{proposition}\label{prop:thetachoice}
	The function
	\begin{equation}
	\label{eq:theta_k}
	\theta_k(\gamma) =\Theta\left ((A_k \gamma)\oslash \mu^k\right)
	\end{equation}
	is continuous and verifies the descent condition (\ref{eq:cond_theta_k}).
\end{proposition}
This construction has several advantages:
\begin{itemize}
\item it allows to choose arbitrarily the parameter $\theta_0$ that will be used eventually when the algorithm is close to convergence (we motivate what are good choices for $\theta_0$ in section \ref{section:local});
\item it is also an easy approach to having an adaptive method, as the approximation of $\Theta_k^*$ has a negligible cost (it only require to solve a one dimensional problem that depends on the smallest value of $(A_k \gamma)\oslash \mu^k$, which can be done in a few iterations of Newton's method).
\item experiments show that this is often an excellent choice in practice.
\end{itemize}

\begin{proof}
Looking at Figure~\ref{phi_omega} can help understanding this proof.	Since the partial derivative of $\partial_\omega \varphi_\omega(x)$ is nonzero for any $x<1$, the implicit function theorem proves the continuity of $\Theta^*$.
	The function $\Theta^*\left((A_k \gamma)\oslash \mu^k)\right)$ is such that every term in relation (\ref{eq:kl_diff_scal}) is non-negative.
	Therefore, by lemma \ref{lemma:F_P_theta}, using this parameter minus $\delta$ ensures the strong decrease (\ref{eq:cond_theta_k}) of the Lyapunov function.
	Constraining the parameter to $[1,\theta_0]$ preserves this properties.% as using $\theta=1$ always satisfies inequality \eqref{eq:cond_theta_k} and using $\theta=\theta_0$ when it is smaller than $\Theta^*\left(\frac{A_k \gamma}{\mu^k}\right)$ does as well.
\end{proof}

The resulting algorithm, which is proved to be convergent by Theorem~\ref{thm:algo}, is written in pseudo-code in Algorithm \ref{SOR}. It uses the function $\Theta$ defined implicitely in \eqref{eq:Theta}, which in practice is approximated with a few iterations of Newton's method on the function $\omega \mapsto \varphi_\omega(\min u)$ which decreasing is as can be seen on Figure~\ref{phi_omega}. Note that with the choice $\theta_0=1$, one recovers exactly the original SK algorithm.

>>>>

\begin{abstract}
This article describes a method for quickly computing the solution to the regularized optimal transport problem. It generalizes and improves upon the widely-used iterative Bregman projections algorithm (or Sinkhorn--Knopp algorithm). 
The idea is to overrelax the Bregman projection operators, allowing for faster convergence. In practice this corresponds to elevating the diagonal scaling factors to a given power, at each step of the algorithm.
We propose a simple method for establishing global convergence by ensuring the decrease of a Lyapunov function at each step.
An adaptive choice of overrelaxation parameter based on the Lyapunov function is constructed.
We also suggest a heuristic to choose a suitable asymptotic overrelaxation parameter, based on a local convergence analysis. Our numerical experiments show a gain in convergence speed by an order of magnitude in certain regimes.
\end{abstract}

@@

The computation of optimal transport between two data relies on the estimation of an optimal transport matrix, the entries of which represent the quantity of mass transported between  data locations. 
Regularization of optimal transport with strictly convex regularization \cite{cuturi13, dessein2016}  nevertheless involves a spreading of the mass. Hence, for particular purposes such as color interpolation \cite{Rabin2014} or gradient flow \cite{2016arXiv160705816C}, it is  necessary  to consider very small regularization of the problem.
In this setting,  the regularized transport problem can be ill-conditioned and the Sinkhorn--Knopp algorithm converges slowly. This is the issue  we want to tackle here.
Before going into further details, we now briefly introduce the main notations and concepts used all along this article.

@@

In \cite{cuturi13}, it has been proposed to regularize this problem by adding a strictly convex entropy regularization:
\begin{equation}\label{ROT}
\min_{\gamma\in\Ccal_1\cap \Ccal_2\cap \IR^{n_1 n_2}_{+}}K^\epsilon(\gamma) \coloneqq \scal{c}{\gamma} 
+ \epsilon \KL(\gamma,\mathbf{1})
,\end{equation}
with $\epsilon>0$, $\mathbf{1}$ is the matrix of size $n_1\times n_2$ full of ones and the Kullback-Leibler divergence is
\begin{equation}\label{KL}
\KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}
\end{equation}
with the convention $0 \log 0= 0$. It was shown in \cite{benamou15}  that the regularized optimal transport matrix $\gamma^*$, which is the unique minimizer of problem \eqref{ROT},  is the Bregman projection of $\gamma^0 = e^{-c/\epsilon}$ (here and in the sequel, exponentiation is meant entry-wise) onto $\Ccal_1 \cap \Ccal_2$:
\begin{equation}\label{eq:reg_ot_pb}
\gamma^* = \argmin_{\Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma)= P_{\Ccal_1 \cap \Ccal_2} (e^{-c/\epsilon}),
\end{equation}
where $P_{\Ccal}$ is the  Bregman projection onto $\Ccal$ defined as
\[
P_{\Ccal}(\xi) \coloneqq \argmin_{\gamma \in \Ccal} \KL(\gamma,\xi).
\]

\subsection{Sinkhorn--Knopp algorithm}
Iterative Bregman projections onto $\Ccal_1$ and $\Ccal_2$ converge to a point in the intersection $\Ccal_1 \cap \Ccal_2$ \cite{bregman67}. Hence, the so-called Sinkhorn--Knopp algorithm (SK) \cite{sinkhorn64} that performs alternate Bregman projections, can be considered to compute the regularized  transport matrix:
\begin{align*}
\gamma^0 &= e^{-c/\epsilon} &
\gamma^{\ell+1} = P_{\Ccal_2}(P_{\Ccal_1}(\gamma^{\ell})),
\end{align*}
and we have 
$\lim_{l\rightarrow +\infty} \gamma^{\ell} = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0) = \gamma^*.$
%
In the discrete setting, these projections correspond to diagonal scalings of the input:
\begin{align}\label{scaling}
P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &\text{with}\quad
a &=  {\mu^1}\oslash{A_1 \gamma} \\
P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &\text{with}\quad
b &= {\mu^2}\oslash{A_2 \gamma}\nonumber
\end{align}
where $\oslash$ is the pointwise division. 
To compute numerically the solution one simply has to store $(a^{\ell}, b^{\ell})\in\IR^{n_1}\times \IR^{n_2}$ and to iterate
\begin{align*}
a^{\ell+1} &= {\mu^1}\oslash{\gamma^0 b^{\ell}} &
b^{\ell+1} &= {\mu^2}\oslash{^t \gamma^0 a^{\ell+1}} .
\end{align*}
We then have $\gamma^{\ell} = \diag(a^{\ell}) \gamma^0 \diag(b^{\ell}).$

Another way to interpret the SK algorithm is as an alternate maximization algorithm on the dual of the regularized optimal transport problem. The dual problem of \eqref{ROT} is
\begin{equation}\label{DROT}
\max_{\substack{\alpha\in \IR^n\\\beta\in \IR^m}}\; E(\alpha,\beta) \coloneqq \langle \alpha,\mu^1\rangle+\langle \beta,\mu^2\rangle-\epsilon\sum_{i,j}e^{(\alpha_i+\beta_j-c_{i,j})/\epsilon}.
\end{equation}
The function $E$ is concave, continuously differentiable and admits a maximizer, so alternate maximization converges and we recover, for $a_i=e^{\alpha_i/\epsilon}$, $b_j=e^{\beta_j/\epsilon}$ and $\gamma^0_{i,j}=e^{-c_{i,j}/\epsilon}$:
\begin{align*}
a^{\ell+1} &= {\mu^1}\oslash{\gamma^0 b^{\ell}} &
b^{\ell+1} &= {\mu^2}\oslash{^t \gamma^0 a^{\ell+1}} 
\end{align*}

@@

As illustrated in Figure \ref{alternate_projections} (a-b), the SK algorithm, that  performs alternate Bregman projections onto the affine sets $\Ccal_1$ and $\Ccal_2$, can be very slow when $\epsilon\to 0$. The idea developed in this paper is to perform overrelaxed projections in order to accelerate the process, as displayed in Figure \ref{alternate_projections} (c).

\begin{figure}[ht!]
\centering
\begin{minipage}[b]{.33\linewidth}
   \centering
   \input{schema_a.tex}
                    \subcaption{}\label{fig:schema_a}
\end{minipage}%
\begin{minipage}[b]{.33\linewidth}
   \centering
   \input{schema_b.tex}
                    \subcaption{}\label{fig:schema_b}
\end{minipage}%
\begin{minipage}[b]{.33\linewidth}
   \centering
   \input{schema_c.tex}
                    \subcaption{}\label{fig:schema_c}
\end{minipage}%
\caption{\label{alternate_projections} The trajectory of $\gamma^{\ell}$ given by the SK algorithm is illustrated for decreasing values of $\epsilon$ in (a) and (b). Overrelaxed projections (c) typically accelerate the convergence rate.}
\end{figure}

@@

We define the $\omega$-overrelaxed projection operator $P^\omega_{\Ccal_k}$ as
\begin{equation}\label{eq:def_or_proj}
\log P^\omega_{\Ccal_k}(\gamma) = (1-\omega) \log \gamma + \omega \log P_{\Ccal_k}(\gamma),
\end{equation}
where the logarithm is taken coordinate-wise.
Note that $P_{\Ccal_k}^0$ is the identity, $P_{\Ccal_k}^1 = P_{\Ccal_k}$ is the standard Bregman projection and $P_{\Ccal_k}^2$ is an involution (in particular because $\Ccal_k$ is an affine subspace).
A naive algorithm would then consist in iteratively applying $P^\omega_{\Ccal_2}\circ P^\omega_{\Ccal_1}$ for some choice of $\omega$.
While it often behaves well in practice, this algorithm may sometimes not converge even for reasonable values of $\omega$.
Our goal in this section is to modify this algorithm to make it robust and to guarantee convergence.

Duality gives another point of view on the iterative overrelaxed Bregman projections. They indeed correspond to a successive overelaxation (SOR) algorithm on the dual objective $E$. This can be seen by using the relationships given after equation~\eqref{DROT}:
\begin{align}\label{SORdual}
\alpha^{\ell+1} &= (1-\omega)\alpha^{\ell} + \omega \arg \max_\alpha E(\alpha,\beta^{\ell})\\
\beta^{\ell + 1}&=(1-\omega) \beta^{\ell} + \omega \arg \max_\beta E(\alpha^{\ell+1},\beta).
\end{align}

\subsection{Lyapunov function}
Convergence of the successive overrelaxed projections is not guaranteed in general. In order to derive a robust algorithm with provable convergence, we introduce the Lyapunov function 
\begin{equation}\label{eq:lyapunov_function}
F(\gamma) = \KL(\gamma^*, \gamma),
\end{equation}
where $\gamma^*$ denotes the solution of the regularized OT problem. We will use this function to enforce the strict descent criterion $F(\gamma^{\ell+1}) < F(\gamma^\ell)$ as long as the process has not converged.
%
The choice of \eqref{eq:lyapunov_function} as a Lyapunov function is of course related to the fact that Bregman projections are used throughout the algorithm. Further, we will show (Lemma \ref{lemma:lyapunov_decrease}) that its decrease is very easy to compute and this descent criterion still allows enough freedom in the choice of the overrelaxation parameter.

@@

@@

\subsection{Proposed algorithm}
We first give a general convergence result that later serves as a basis to design an explicit algorithm.
\begin{theorem}\label{thm:algo}
Let $\theta_1$ and $\theta_2$ be two continuous functions of $\gamma$ such that
	\begin{equation}\label{eq:cond_theta_k}
	\forall \gamma \in \IR_{+*}^{n_1 n_2},\quad
	F(P_{\Ccal_k}^{\theta_k(\gamma)}(\gamma)) \le F(\gamma) ,
	\end{equation}
	where the inequality is strict whenever $\gamma \notin \Ccal_k$.
	Consider the sequence defined by $\gamma^0 = e^{-c/\epsilon}$ and
	\begin{align*}
	\tilde{\gamma}^{\ell+1} &= P_{\Ccal_1}^{\theta_1(\gamma^{\ell})}(\gamma^{\ell}) \\
	\gamma^{\ell+1} &= P_{\Ccal_2}^{\theta_2(\tilde{\gamma}^{\ell+1})}(\tilde{\gamma}^{\ell+1}).
	\end{align*}
	Then the sequence $(\gamma^{\ell})$ converges to $\gamma^*$.
\end{theorem}

\begin{lemma}[]
	\label{lemma:trivial_intersection}
	Let us take $\gamma^0$ in $\IR_{+*}^{n_1 n_2}$,
	and denote
	\[
	S = \left\{
	\diag(a) \gamma^0 \diag(b),\quad
	(a,b) \in \IR_{+*}^{n_1 + n_2}
	\right\}
	\]
	the set of matrices that are diagonally similar to $\gamma^0$.
	Then the set $S \cap \Ccal_1 \cap \Ccal_2$ contains exactly one element $\gamma^* = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0)$.
\end{lemma}
\begin{proof}
	We refer to \cite{cuturi13} for a proof of this lemma.
\end{proof}

\begin{proof}[Proof of the theorem]
	First of all, notice that the operators $P_{\Ccal_k}^\theta$ apply a scaling to lines or columns of matrices. All $(\gamma^{\ell})$ are thus diagonally similar to $\gamma^0$:
	\[
	\forall \ell\ge0,\quad \gamma^{\ell} \in S
	\]
	By construction of the functions $\theta_k$, the sequence of values of the Lyapunov function $(F(\gamma^{\ell}))$ is non-increasing. Hence $(\gamma^{\ell})$ is precompact.
	%
	If $\xi$ is a cluster point of $(\gamma^{\ell})$, let us define
	\begin{align*}
	\tilde{\xi} &= P_{\Ccal_1}^{\theta_1(\xi)}(\xi) \\
	\xi' &= P_{\Ccal_2}^{\theta_2(\tilde{\xi})}(\tilde{\xi}).
	\end{align*}
	Then by continuity of the applications, $F(\xi) = F(\tilde{\xi}) = F(\xi')$.
	From the hypothesis made on $\theta_1$ and $\theta_2$, it can be deduced that $\xi$ is in $\Ccal_1$ and that $\tilde{\xi}$ is in $\Ccal_2$. Therefore $\xi' = \tilde{\xi} = \xi$ is in the intersection $\Ccal_1 \cap \Ccal_2$.
	By Lemma \ref{lemma:trivial_intersection}, $\xi = \gamma^*$, and the whole sequence $(\gamma^{\ell})$ converges to the solution.
\end{proof}

@@

To construct explicitly functions $\theta_k$ as needed in Theorem~\ref{thm:algo}, we need the following lemma.
\begin{lemma}\label{lemma:F_P_theta}
	Let $1\le \theta < \omega$. Then, for any $\gamma \in \IR_{+*}^{n_1 n_2}$, one has
	\begin{equation}\label{eq:F_P_theta}
	F(P^\theta_{\Ccal_k}(\gamma)) \le F(P^\omega_{\Ccal_k}(\gamma)).
	\end{equation}
	Moreover, equality occurs if and only if $\gamma \in \Ccal_k$.
\end{lemma}
\begin{proof}
	Thanks to lemma \ref{lemma:lyapunov_decrease}, one knows that
	\[
	F(P^\theta_{\Ccal_k}(\gamma)) - F(P^\omega_{\Ccal_k}(\gamma))
	= \scal{\mu^k}{(\varphi_\omega - \varphi_\theta) \left( \frac{A_k \gamma}{\mu^k} \right) } .
	\]
	The function that maps $t \in [1,\infty)$ to $\varphi_t(x)$ is non-increasing since
	$\partial_t \varphi_t(x) =  (x^{1-t} - 1)\log x.$
	For $x\neq 1$, it is even strictly decreasing.
	Thus inequality \eqref{eq:F_P_theta} is valid, with equality \emph{iff} $A_k \gamma = \mu$.
\end{proof}

@@

\begin{proposition}\label{prop:thetachoice}
	The function
	\begin{equation}
	\label{eq:theta_k}
	\theta_k(\gamma) =\Theta\left ((A_k \gamma)\oslash \mu^k\right)
	\end{equation}
	is continuous and verifies the descent condition (\ref{eq:cond_theta_k}).
\end{proposition}
This construction has several advantages:
\begin{itemize}
\item it allows to choose arbitrarily the parameter $\theta_0$ that will be used eventually when the algorithm is close to convergence (we motivate what are good choices for $\theta_0$ in section \ref{section:local});
\item it is also an easy approach to having an adaptive method, as the approximation of $\Theta^*$ has a negligible cost (it only require to solve a one dimensional problem that depends on the smallest value of $(A_k \gamma)\oslash \mu^k$, which can be done in a few iterations of Newton's method).
\item experiments show that this is often an excellent choice in practice.
\end{itemize}

\begin{proof}
Looking at Figure~\ref{phi_omega} can help understand this proof.	Since the partial derivative of $\partial_\omega \varphi_\omega(x)$ is nonzero for any $x<1$, the implicit function theorem proves the continuity of $\Theta^*$.
	The function $\Theta^*\left((A_k \gamma)\oslash \mu^k)\right)$ is such that every term in relation (\ref{eq:kl_diff_scal}) is non-negative.
	Therefore, by lemma \ref{lemma:F_P_theta}, using this parameter minus $\delta$ ensures the strong decrease (\ref{eq:cond_theta_k}) of the Lyapunov function.
	Constraining the parameter to $[1,\theta_0]$ preserves this property.% as using $\theta=1$ always satisfies inequality \eqref{eq:cond_theta_k} and using $\theta=\theta_0$ when it is smaller than $\Theta^*\left(\frac{A_k \gamma}{\mu^k}\right)$ does as well.
\end{proof}

The resulting algorithm, which is proved to be convergent by Theorem~\ref{thm:algo}, is written in pseudo-code in Algorithm \ref{SOR}. It uses the function $\Theta$ defined implicitly in \eqref{eq:Theta}, which in practice is approximated with a few iterations of Newton's method on the function $\omega \mapsto \varphi_\omega(\min u)$ which decreasing is as can be seen on Figure~\ref{phi_omega}. Note that with the choice $\theta_0=1$, one recovers exactly the original SK algorithm.
